Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-06-06 GWLDCI.tex
continuing minor edits to paper 1

\note{sian note to self: editing has reached this point}

@@

We prove the following claim: $W'$ is exact if and only if $V(P)$ contains no circuits $C$ with $\rk C< |P|$ in $(P, V(P))$. Since $\rk(M(W'))$ is bounded above by $|P|$, the result follows.

Suppose $C \subseteq V(P)$ is a circuit of rank $m < |P|$. By Lemma \ref{lem facts about WLD matroids}, we know that $|\Prop_{W'}(C)| = m$ as well, where the subscript to $\Prop$ specifies the diagram we are working in.  Observe that $\Prop_{W'}(C)\subseteq P$ by definition. The set $P \setminus \Prop_{W'}(C)$ is thus nonempty, and we can consider the subdiagram $W'':= (P\setminus \Prop_{W'}(C),V(P\setminus\Prop_{W'}(C))$. By the density condition on subdiagrams of admissible diagrams, we have
\[|V(P\setminus\Prop_{W'}(C))| \geq |P\setminus\Prop_{W'}(C)| + 3.\]
It is easy to verify that $V(P\setminus\Prop_{W'}(C)) \subseteq V(P)\setminus C$; since $C \subseteq V(P)$ and $\Prop_{W'}(C) \subseteq P$, we can therefore rewrite the previous inequality as
\[|V(P)| - (m+1) \geq |V(P\setminus\Prop(C))| \geq |P| - m + 3.\]
Simplifying, we obtain $|V(P)| \geq |P| + 4$, i.e. $(P,V(P))$ is not an exact diagram.

Conversely, suppose that $W'$ is not exact and for a contradiction suppose also that $M(W')$ is uniform of rank $|P|$.  Take $p \in P$.  Then $|V(P)\setminus V(p)| = |V(P)| - 4 \geq |P|$ by non-exactness.  By uniformity there is an independent set of size $|P|$ in $V(P)\setminus V(p)$.  This is impossible because the submatrix corresponding to this independent set has $|P|$ rows but the one corresponding to $p$ is all $0$ one of them is all 0 so it cannot be full rank.
%suppose that $(P,V(P))$ is not exact, i.e. $|V(P)| \geq |P| +4$. We have 
%\[|V(P)| - (m+1) \geq |P| - m + 3\]
%as above, but \hlfix{we need to know that $P\setminus\Prop(C)$ is non-empty before we can complete the argument.}{and if we knew this, it would be immediate that $\rk(C) < |P|$}
\end{proof}

@@

\begin{dfn}
Given a Wilson loop diagram $W = (\cP, [n])$, define the positroid cell associated to a Wilson loop diagram, $\Sigma(W)$, to be the cell in the CW complex on $\Gr(|\cP|, n)$ defined by the positroid $M(W)$.
\end{dfn}

@@

Now we are ready to prove the main results of this section, namely that two Wilson loop diagrams define the same matroid if and only if they are equivalent (Theorem \ref{same matroid iff equiv}).
We also
give a formula for the number of Wilson loop diagrams in a given equivalence class (Corollary \ref{number of equiv diagrams}), completing the characterization of the correspondence between Wilson Loop diagrams and matroids starting in \cite{wilsonloop}.

@@

In this section we have characterized the correspondence between Wilson loop diagrams and positroids, showing that Wilson Loop diagrams define the same matroid if and only if they are equivalent (Theorem \ref{same matroid iff equiv}). Corollary \ref{number of equiv diagrams} enumerates the fibres of the map from Wilson Loop diagrams to positroids.  The obvious next question would be to enumerate the number of equivlence classes.  We do not obtain a formula for this, but do obtain an alternate characterization of the equvalence classes which is the subject of the next section.

@@

A \emph{polytope} can equivalently be defined as the convex hull of a finite set of points in some $\mathbb{R}^d$ or as the intersection of finitely many closed half spaces in $\mathbb{R}^d$ with the additional property that this intersection is bounded in the sense of not containing any ray (see chapter 0 of \cite{Ziegler}).

A \emph{face} of a polytope is the intersection of the polytope with any hyperplane for which the entire polytope is in one of the closed half-spaces defined by the hyperplane.  The polytope itself is also considered a face (in the case that that polytope is of full dimension, this definition can be unified by considering linear equalities (including the trivial one) in place of hyperplanes and the corresponding linear inequalities in place of the half-spaces defined by the hyperplane, see chapter 2 of \cite{Ziegler}).  Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.  In particular parallel faces must be of the same dimension and any two vertices are parallel to each other.

There are many important polytopes in combinatorics, including, of primary importance to us, the associahedron.

Two polytopes are \emph{combinatorially equivalent} if there is a bijection between their faces which preserves inclusion of faces.  Often in combinatorics one is primarily interested in the combinatorial equivalence classes, and common names, including the associahedron, refer to these equivalence classes.  A particular polytope in the class is then known as a \emph{realization}.

Associahedra can also be viewed in terms of polygon dissections.  Fix $n$ and consider dissections of a convex $n$-gon.  The $n-1$ asociahedron can be given in terms of the dissections of this $n$-gon; the vertices of the associahedron are the triangulations of the $n$-gon while the dimension $k$ faces are the dissections which are $k$ diagonals away from being a triangulation.  One face is included in another if the first dissection is a refinement of the second.

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}; the following will be the one which we will use.  See Figure~\ref{associahedron5gon} for examples.

\begin{dfn}
  Let $x_1, \ldots, x_n$ define a convex $n$-gon.  Let $T$ be the set of triangulations of this $n$-gon.  For each $t\in T$ define $s_t$ to be the point in $\mathbb{R}^n$ whose $i$th coordinate is the sum of the areas of all triangles of $t$ which are incident to $x_i$.  Let $A_n$ be the convex hull of the $s_t$.
\end{dfn}
This is the construction of the secondary polytope (so $A_n$ is the secondary polytope of $x_1, \ldots, x_n$), see \cite{Ziegler} Definition 9.9.  In general the $x_1, \ldots, x_n$ may be any points in any $\mathbb{R}^d$.  Triangulations in this more general context are decompositions into simplices and the area of the triangles is replaced by the volume of the simplices.

$A_n$ defined in this way is a realization of the $n-1$ associahedron (\cite{Ziegler}, Example 9.11).  Note that we get a realization for any convex $n$-gon, but these realizations are all equally good for our purposes, and so we fix $x_1, \ldots, x_n$ defining a particular convex $n$-gon from now on.

@@

\begin{figure}
\includegraphics{associahedron5gon} \hspace{1cm} \includegraphics{associahedron3d}
\caption{Two associahedra labelled by all and some (respectively) of their polygon dissections.}\label{associahedron5gon}
\end{figure}

Note that Wilson loop diagrams which are weakly admissible but not admissible are necessarily exact and hence have $\tau$ graphs which are triangulations, hence dimension 0 faces (vertices) in $A_n$.  All of these are equivalent in the sense that they are only retriangulations away from each other.  It will be convenient in what follows, to include these diagrams.  The next two propositions would be identical in the case of admissible instead of weakly admissible, while the corollary could be changed to involve admissible diagrams only by subtracting one from the number of non parallel faces in order to account for the missing dimension 0 faces.

@@

\begin{proof}
  Let $W$ and $X$ be two equivalent Wilson loop diagrams.  From Corrolary \ref{equialentretriangulation} we know that $\tau(W)$ and $\tau(X)$ differ from each other only by retriangulations of their triangulated pieces.

@@

  Consider now an edge whose quadrilateral has corners $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From the definition of $A_n$ we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$, and all of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates.  This is illustrated in Figure~\ref{triangulate}.

@@

    Since $\tau(W)$ and $\tau(X)$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $\tau(W)$ that crosses through a non-triangle face $f$ of $\tau(X)$. 
    It will be most convenient to look at normal vectors and to use the choice of normal vector discussed on p8 of \cite{CSZinequivalent}, namely pick one side of $d$ to be positive and define the $i$th coordinate of the $\omega^+_d$ to be
    \[
    \omega^+_d(i) = \begin{cases} \text{dist}(x_i, d) & \text{if $x$ is on the positive side of $d$}\\ 0 & \text{otherwise} \end{cases}
    \]
    We have from \cite{CSZinequivalent} that $\omega^+_d$ is a normal vector for any face with $d$ as a diagonal, however we can also check this fact in a completely elementary way using the shoelace formula for the area of a triangle.  The shoelace formula says that if $(x_1, y_1), (x_2, y_2), (x_3, y_3)$ are the coordinates of the corners of a triangle in counterclockwise order then the area of the triangle is $(x_1y_2 - x_1y_3 + x_2y_3 - x_2y_1 + x_3y_1 - x_3y_2)/2$.

    Proceed now to check that $\omega^+_d$ is a normal vector for any face associated to a dissection including $d$.  We will do this by showing it is normal to every edge bounding the face.  To do so, choose the coordinate system for the polygon so that $d$ is along the $x$-axis.  Let $(a_i, h_i)$, $i=1,2,3,4$ be the coordinates of the vertices of a quadrilateral in counterclockwise order and which corresponds to an edge bounding the face.  Since the dissection includes $d$ either all four points are on the negative side of $d$, and so the dot product of $\omega^+_d$ with the direction vector for the quadrilateral is immediately $0$, or, by the formula in the proof of the previous proposition and the shoelace formula, the dot product of $\omega^+_d$ with the vector associated to the quadrilateral is
    \begin{align*}
      & h_1(a_2h_3-a_2h_4+a_3h_4-a_3h_2+a_4h_2-a_4h_3)/2 - h_2(a_1h_3-a_1h_4+a_3h_4-a_3h_1+a_4h_1-a_4h_3)/2 \\
      & + h_3(a_1h_2-a_1h_4+a_2h_4-a_2h_1+a_4h_1-a_4h_2)/2 - h_4(a_1h_2-a_1h_3+a_2h_3-a_2h_1+a_3h_1-a_3h_2)/2 \\
      & = 0
    \end{align*}

@@

The two propositions imply
\begin{cor}
  The number of inequivalent weakly admissible Wilson loop diagrams on $n$ is the number of non-parallel faces in the associahedron $A_n$.
\end{cor}

>>>>

@@

We prove the following claim: $W'$ is exact if and only if $V(P)$ contains no circuits $C$ with $\rk (C)< |P|$ in $(P, V(P))$. Since $\rk(M(W'))$ is bounded above by $|P|$, the result follows.

First suppose $C \subseteq V(P)$ is a circuit of rank $m < |P|$. Observe that $\Prop_{W'}(C)\subseteq P$ by definition, where the subscript to $\Prop$ specifies the diagram we are working in. By Lemma \ref{lem facts about WLD matroids} we know that ${|\Prop_{W'}(C)| = m}$; thus the set $P \setminus \Prop_{W'}(C)$ must be nonempty, and we can consider the subdiagram $W'':= (P\setminus \Prop_{W'}(C),V(P\setminus\Prop_{W'}(C))$. By the density condition on subdiagrams of admissible diagrams, we have
\[|V(P\setminus\Prop_{W'}(C))| \geq |P\setminus\Prop_{W'}(C)| + 3.\]
It is easy to verify that $V(P\setminus\Prop_{W'}(C)) \subseteq V(P)\setminus C$. Since $C \subseteq V(P)$ and $\Prop_{W'}(C) \subseteq P$, it follows from the previous inequality that
\[|V(P)| - (m+1) \geq |V(P\setminus\Prop(C))| \geq |P| - m + 3.\]
Simplifying, we obtain $|V(P)| \geq |P| + 4$, i.e. $(P,V(P))$ is not an exact diagram.

Conversely, suppose that $W'$ is not exact and for a contradiction suppose also that $M(W')$ is uniform of rank $|P|$.  Take $p \in P$.  Then $|V(P)\setminus V(p)| = |V(P)| - 4 \geq |P|$ by non-exactness.  By uniformity there must be an independent set of size $|P|$ in $V(P)\setminus V(p)$.  However, this is impossible because the submatrix of $C(W')$ corresponding to this independent set has $|P|$ rows but the one corresponding to $p$ is all $0$, so it cannot be full rank.
%suppose that $(P,V(P))$ is not exact, i.e. $|V(P)| \geq |P| +4$. We have 
%\[|V(P)| - (m+1) \geq |P| - m + 3\]
%as above, but \hlfix{we need to know that $P\setminus\Prop(C)$ is non-empty before we can complete the argument.}{and if we knew this, it would be immediate that $\rk(C) < |P|$}
\end{proof}

@@

\begin{dfn}
Given a Wilson loop diagram $W = (\cP, [n])$ and its associated positroid $M(W)$, let $\Sigma(W)$ denote the corresponding positroid cell in the CW complex on $\Gr(|\cP|, n)$.
\end{dfn}

@@

Now we are ready to prove the main results of this section, namely that two Wilson loop diagrams define the same matroid if and only if they are equivalent (Theorem \ref{same matroid iff equiv}).
We also
give a formula for the number of Wilson loop diagrams in a given equivalence class (Corollary \ref{number of equiv diagrams}), completing the characterization of the correspondence between Wilson loop diagrams and matroids starting in \cite{wilsonloop}.

@@

The obvious next step would be to enumerate the number of equivalence classes.  We do not obtain a formula for this, but in the next section we give an alternate characterization of the equivalence classes in terms of faces in associahedra.

@@

We first give a quick summary of the polytope terminology we will use in this section; the interested reader is referred to \cite{Ziegler} for more details.

A \emph{polytope} can equivalently be defined as the convex hull of a finite set of points in some $\mathbb{R}^d$, or as the intersection of finitely many closed half spaces in $\mathbb{R}^d$ with the additional property that this intersection is bounded in the sense of not containing any ray (see chapter 0 of \cite{Ziegler}).

A \emph{face} of a polytope is the intersection of the polytope with any hyperplane for which the entire polytope is in one of the closed half-spaces defined by the hyperplane.  The polytope itself is also considered a face; in the case that that polytope is of full dimension, this definition can be unified by considering linear equalities (including the trivial one) in place of hyperplanes and the corresponding linear inequalities in place of the half-spaces defined by the hyperplane (see chapter 2 of \cite{Ziegler}).

Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.  In particular parallel faces must be of the same dimension and any two vertices are parallel to each other.

Finally, two polytopes are \emph{combinatorially equivalent} if there is a bijection between their faces which preserves inclusion of faces.  Often in combinatorics one is primarily interested in the combinatorial equivalence classes, and common names, including the associahedron, refer to these equivalence classes.  A particular polytope in the class is then known as a \emph{realization}.

There are many important polytopes in combinatorics, including, of primary importance to us, the associahedron. The associahedron can be characterized in many different ways; here is one definition.

\begin{dfn}\label{def:associahedron}
Fix $n$ and consider dissections of a convex $n$-gon.  The $n-1$ {\em associahedron} can be given in terms of the dissections of this $n$-gon; the vertices of the associahedron are the triangulations of the $n$-gon while the $k$-dimensional faces are the dissections which are $k$ diagonals away from being a triangulation.  One face is included in another if the first dissection is a refinement of the second. 
\end{dfn}

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}; the following realization will be the one which we will use.

\begin{dfn}\label{def:secondary polytope}
  Let $x_1, \ldots, x_n$ define a convex $n$-gon, and let $T$ be the set of triangulations of this $n$-gon.  For each $t\in T$ define $s_t$ to be the point in $\mathbb{R}^n$ whose $i$th coordinate is the sum of the areas of all triangles of $t$ which are incident to $x_i$.  Let $A_n$ be the convex hull of the $s_t$; then $A_n$ a realization of the $n-1$ associahedron (\cite{Ziegler}, Example 9.11)
\end{dfn}
Note that this construction works for any convex $n$-gon, but these realizations are all equally good for our purposes, and so we fix $x_1, \ldots, x_n$ defining a particular convex $n$-gon from now on.

Definition~\ref{def:secondary polytope} is a special case of the {\em secondary polytope} construction; see \cite[Definition 9.9]{Ziegler} for more details.  In general the $x_1, \ldots, x_n$ may be any points in any $\mathbb{R}^d$.  Triangulations in this more general context are decompositions into simplices and the area of the triangles is replaced by the volume of the simplices.

@@

\begin{figure}
\includegraphics{associahedron5gon} \hspace{1cm} \includegraphics{associahedron3d}
\caption{Two associahedra, labelled by all and some (respectively) of their polygon dissections.}\label{associahedron5gon}
\end{figure}

Note that Wilson loop diagrams which are weakly admissible but not admissible are necessarily exact, and hence have $\tau$ graphs which are triangulations, i.e. dimension 0 faces (vertices) in $A_n$.  All of these are equivalent (as Wilson loop diagrams) in the sense that they are only retriangulations away from each other; meanwhile, as noted above, all vertices of $A_n$ are parallel to each other.  It will be convenient in what follows to include these diagrams, but the results are easily modified to remove them if needed: Propositions~\ref{prop easy way} and \ref{prop hard way} would be identical if we replaced ``weakly admissible'' by ``admissible'', while Corollary~\ref{cor:count inequiv diagrams} could be changed to consider only admissible diagrams by subtracting 1 from the number of non-parallel faces in order to account for the missing dimension 0 faces.

@@

\begin{proof}
  Let $W$ and $X$ be two equivalent Wilson loop diagrams.  From Corollary \ref{equialentretriangulation} we know that $\tau(W)$ and $\tau(X)$ differ from each other only by retriangulations of their triangulated pieces.

@@

  Consider now an edge whose quadrilateral has corners $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From the definition of $A_n$ we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$, and all of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is 
  \[(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))\]
  on $i$, $j$, $k$, $l$ and $0$ in all other coordinates.  This is illustrated in Figure~\ref{triangulate}.

@@

    Since $\tau(W)$ and $\tau(X)$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $\tau(W)$ that crosses through a non-triangle face $f$ of $\tau(X)$. 
    It will be most convenient to look at normal vectors and to use the choice of normal vector discussed on p8 of \cite{CSZinequivalent}, namely pick one side of $d$ to be positive and define the $i$th coordinate of the $\omega^+_d$ to be
    \[
    \omega^+_d(i) = \begin{cases} \text{dist}(x_i, d) & \text{if $x$ is on the positive side of $d$}\\ 0 & \text{otherwise} \end{cases}
    \]
    We have from \cite{CSZinequivalent} that $\omega^+_d$ is a normal vector for any face with $d$ as a diagonal, however we can also check this fact in a completely elementary way using the shoelace formula for the area of a triangle.  The shoelace formula says that if $(x_1, y_1), (x_2, y_2), (x_3, y_3)$ are the coordinates of the corners of a triangle in counterclockwise order then the area of the triangle is 
    \[(x_1y_2 - x_1y_3 + x_2y_3 - x_2y_1 + x_3y_1 - x_3y_2)/2.\]

    Proceed now to check that $\omega^+_d$ is a normal vector for any face associated to a dissection including $d$.  We will do this by showing it is normal to every edge bounding the face.  To do so, choose the coordinate system for the polygon so that $d$ is along the $x$-axis.  Let $(a_i, h_i)$, $i=1,2,3,4$ be the coordinates of the vertices of a quadrilateral in counterclockwise order and which corresponds to an edge bounding the face.  Since the dissection includes $d$ either all four points are on the negative side of $d$, and so the dot product of $\omega^+_d$ with the direction vector for the quadrilateral is immediately $0$, or, by the formula in the proof of the previous proposition and the shoelace formula, the dot product of $\omega^+_d$ with the vector associated to the quadrilateral is
    \begin{align*}
       & \hspace{1cm}\textstyle\frac{1}{2}h_1(a_2h_3-a_2h_4+a_3h_4-a_3h_2+a_4h_2-a_4h_3) \\
       &\hspace{3cm} - \textstyle\frac{1}{2} h_2(a_1h_3-a_1h_4+a_3h_4-a_3h_1+a_4h_1-a_4h_3) \\
       &\hspace{5cm}+ \textstyle\frac{1}{2}h_3(a_1h_2-a_1h_4+a_2h_4-a_2h_1+a_4h_1-a_4h_2) \\
       &\hspace{7cm}- \textstyle\frac{1}{2}h_4(a_1h_2-a_1h_3+a_2h_3-a_2h_1+a_3h_1-a_3h_2), 
    \end{align*}
    which evaluates to 0.

@@

Propositions~\ref{prop easy way} and \ref{prop hard way} together imply:
\begin{cor}\label{cor:count inequiv diagrams}
  The number of inequivalent weakly admissible Wilson loop diagrams on $n$ is the number of non-parallel faces in the associahedron $A_n$.
\end{cor}
